Customizing Software by Direct Manipulation of Structured Data

### 2020-05-21 paper.tex
dnj edits

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we propose \emph{table-driven customization,} a new
  paradigm for enabling end users to customize software applications
  without doing traditional programming\emph{.} Users directly
  manipulate a tabular view of the structured data inside the
  application, rather than writing imperative scripts as in most
  customization tools. We extend this simple model with a spreadsheet
  formula language and custom data editing widgets, which provide
  sufficient expressivity to implement many useful customizations.

  We describe Wildcard, a browser extension which implements
  table-driven customization in the context of web applications. Through
  concrete examples, we demonstrate that this paradigm can be used to
  create useful customizations for real applications. We share
  reflections from our usage of the Wildcard system, including its
  strengths and limitations relative to other customization approaches.
  We further explore how new software architectures might encourage
  application developers to promote this style of end-user
  customization.
\end{abstract}

@@

There have been many attempts at empowering end users to customize their
software by offering them a simplified programming tool. Some scripting
languages (AppleScript, Chickenfoot) have a friendly syntax that
resembles natural language. Visual programming tools (Mac Automator,
Zapier) eliminate text syntax entirely. Macro recorders (Applescript,
Helena, WebVCR) remove some of the initial programming burden by letting
a user start by concrete demonstrations.

These approaches have many differences, but they all share something in
common: an imperative programming model, with mutable variables,
conditionals and loops. End users ultimately must use these traditional
programming constructs to express their ideas, and the object of
interest is a \emph{script}, a sequence of commands.

We have known for decades about an alternate approach: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the de facto standard in GUIs today, but when it comes
to customizing software, it is rarely to be found. In this work, we ask:
what would it look like to build a software customization interface that
relies on direct manipulation? We take inspiration from spreadsheets and
visual database query interfaces \citep{2020a, bakke2016}, which have
successfully enabled end users to run queries and computations through
direct manipulation of data.

In this paper we present a technique called \emph{table-driven
customization} which applies these ideas from visual query interfaces in
the context of software customization. We augment an application's UI
with a table view, where the user can see and manipulate the structured
data inside an application. Changes in the table view result in
immediate corresponding changes to the original user interface of the
application, enabling the user to customize an application with live
feedback.

We have developed a browser extension called Wildcard which uses web
scraping techniques to implement table-driven customization for existing
Web applications. In Section~\ref{sec:examples} we concretely introduce
the key ideas of table-driven customization by presenting several
examples of real customizations implemented in Wildcard.

In Section~\ref{sec:architecture} we explain the architecture of
table-driven customization. We focus on the \emph{table adapter}
abstraction, which allows many different types of underlying state to be
bidirectionally mapped to a table of data. We describe certain types of
table adapters we've built in Wildcard; we also describe future adapters
that could be supported in the general paradigm.

We have used Wildcard to build real customizations for 11 different
websites. In Section~\ref{sec:evaluation} we present reflections from
this process: customizations we were able to build, limitations we
encountered, and reflections on the ease of integrating scraping logic
with real websites.

@@

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: End users should be able
  to customize an application by examining and modifying its data,
  rather than by writing imperative scripts.
\item
  \emph{Third-party semantic wrappers}: Typically, tools that don't rely
  on official extension APIs resort to offering low-level APIs for
  customization. Instead, we suggest a community-maintained library of
  semantic wrappers around existing applications, enabling end users to
  work with domain objects rather than low-level representations.
\end{itemize}

@@

First, the three different data sources are each bidirectionally mapped
to a table interface by a \textbf{table adapter}. The table adapter
defines how to map a particular type of data to a table, and what
effects edits should have on the original data source. In some cases,
the mapping logic is straightforward: the local storage adapter stores a
table of data, so the mapping to the table abstraction is trivial. In
other cases, the mapping is more involved: the DOM scraping adapter
implements web scraping logic to produce a table of data from the web
page, and edits to the table result in DOM manipulations like reordering
rows of data on the page.

Ultimately, we provide the user with a single combined table view that
they can edit, rather than three separate tables. The \textbf{query
engine} is responsible for combining the tables of data into this single
view, and routing the user's edits back to the individual table
adapters. In this example, the query engine has joined the three tables
together by a shared ID column, and sorted the result by the name
column.

@@

A key idea in table-driven customization is that different data sources
can all be mapped to the generic abstraction of an editable table. In a
relational database, the table matches the underlying storage format,
but in table-driven customization, the table is merely an
\emph{interface layer}. The data shown in the table is a projection of
some underlying state, and edits to the table can have arbitrary effects
on the underlying state.

Externally, a table adapter must satisfy an abstract interface. The
first two parts of the interface largely resemble the interface exposed
by a table in a relational database:

\textbf{Returning a table}: A table adapter exposes a table of data: an
ordered list of records. Each record contains associates attributes and
values. Tables have a typed schema, so the same attributes are shared
across all records. A table adapter can update the contents of a table
at any time, in response to changes in the underlying state (e.g., a DOM
scraping adapter can update the table when the page body changes). New
data is pushed out to other components of the system, and the query view
reactively updates in response.

\textbf{Making edits}: The query engine can request to a table adapter
to make an edit to a record. The meaning of making an edit can vary
depending on the adapter: in the local storage adapter, an annotation
can be persisted into local storage; in the DOM scraping adapter, an
edit can represent filling in a form field. An adapter can also mark
values as read-only if it wouldn't be meaningful to edit them; for
example, the DOM scraping adapter typically marks page content as
read-only, except for editable form fields.

@@

Here we describe in more detail the table adapters which we have
implemented in Wildcard to power the customizations shown in
Section~\ref{sec:examples}.

\textbf{DOM scraping adapters} are the essential component that enables
Wildcard to interface with an existing website UI. In addition to the
standard web scraping problem of extracting a table of data from the
DOM, a scraping adapter must also manipulate the DOM to re-order rows,
edit form entries, and inject annotations as the table is edited.

Because each website has unique content, we rely on programmers to
create a DOM scraping adapter for each individual website to make it
available for customization in Wildcard. To make this approach viable,
we have built library functions that make it as easy as possible to
implement a scraping adapter, only requiring the programmer to implement
the minimal site-specific parts. The programmer writes a single scraping
function which uses CSS selectors and DOM APIs to extract and return
relevant elements from the page. The library functions then wrap this
function to implement the rest of the needed functionality; for example,
when the table is sorted, we find the DOM elements corresponding to the
rows in the table, remove them from the page, and then reinsert them in
the new order.

An \textbf{AJAX scraping adapter} intercepts AJAX requests made by a web
page, and extracts information from those requests. When available, this
tends to be a helpful technique because the data is already in a
structured form, and often includes information not shown in the UI. As
with DOM scraping adapters, we have made it as easy as possible for
programmers to create site-specific AJAX scraping adapters. A programmer
writes a function which specifies how to extract data from an AJAX
request, and the framework handles the details of actually intercepting
requests and calling the programmer-defined function.\footnote{So far we
  have only implemented AJAX scraping in the Firefox version of
  Wildcard, since Firefox has convenient APIs for intercepting requests.
  It appears possible to implement in Chrome as well, but we have not
  finished our implementation.}

The \textbf{local storage adapter} simply stores a table of data in the
browser. As shown in the example use cases, this can be useful for
privately persisting annotations, without needing to upload them to a
web service.

@@

We have designed the table adapter API to be general enough to support
other types of useful adapters in the future. Here are three examples of
future possibilities:

\textbf{Integrated website adapters}: A key benefit of the table adapter
abstraction is that Wildcard is not coupled to web scraping as the only
means for integrating with existing sites, but can also accommodate
first party developers adding support directly into their own websites.
A DOM Scraping table adapter could be swapped out for an ``integrated
website adapter'' which directly accesses the internal state of the
application, without needing to extract it from the UI.

With the advent of rich frontend web frameworks, structured application
state is now often available in the web client. We suspect it is
possible to create plugins for frontend frameworks which expose
structured state to Wildcard with minimal effort from the application
developers. For example, we have created an initial prototype of a
plugin for the Redux state management library, which represents the
state of a user interface as a single object. To configure an integrated
website adapter for such an existing application, the user simply
specifies two functions. The first one projects the centralized
application state into a table, and define handlers for how data edits
to the table should affect the state.

\textbf{Shared storage adapter}: It would be useful to share user
annotations between people and across devices---for example,
collaboratively taking notes with friends on a list of options for
places to stay on Airbnb. The existing Local Storage Adapter could be
extended to share live synchronized data with other users. This could be
achieved through a centralized web server, or through P2P connections
which might provide stronger privacy guarantees.

\textbf{Third party API data adapter}: Currently, the main mechanism for
including data from web APIs in Wildcard is using spreadsheet formulas.
However, a web API could also be wrapped to expose a table API that
would dynamically create tables in response to queries. For example,
when fetching walkability scores for many GPS locations, the query
engine could request a table of walkability scores for various pairs of
latitude and longitude, and the table adapter could dynamically perform
API queries to populate a result table. (\emph{todo: make a stronger
case for why this is useful or better than formulas})

(\emph{note: another ordering for the adapters could be by function.
First, DOM, AJAX, and Redux, which all integrate with the original app.
Then, Local Storage, Shared Storage, API, which are all about adding
functionality to the site. This might make more logical sense, but it
risks confusing what's already been done and what's a future
possibility. The current order seems good enough?})

@@

First, every query involves a primary DOM scraping table adapter which
associates records in the result with elements in the application's user
interface. At minimum, the primary table adapter needs to return record
IDs and have the ability to manipulate the application's UI. It can also
optionally return data about each record.

Next, additional tables (AJAX data, local storage data) are left joined
by ID. (\emph{todo: discuss IDs here?}) Finally, the result table can be
sorted and filtered by any column.

One way to think of this model is a tiny constrained subset of the SQL
query model. We've found that this simple model has proven sufficient
for meeting the needs of customization in practice, and minimizes the
complexity of supporting more general and arbitrary queries. But because
it fits into the general SQL paradigm, it could theoretically be
extended to support more types of queries.

The query engine is also responsible for executing formulas. We have
built a small language resembling the formula languages in visual query
tools like SIEUFERD. As in those tools, and unlike in spreadsheets,
formulas automatically apply to an entire column of data, and can
reference columns rather than only specific cells.

@@

Ainsworth et al's provide a helpful taxonomy of the value of multiple
representations \citep{ainsworth1999}. In their terms, Wildcard plays a
\emph{complementary role} by supporting a different set of tasks from
the original application, while displaying \emph{shared information}. We
also suspect that Wildcard helps \emph{constructs deeper understanding}
by \emph{subtraction}. By stripping away details and only showing the
essential data in an interface, Wildcard helps users think of new ways
of using that data, outside the specific constraints of the original
application.

It is important to note that there are many specific customizations that
can be achieved in systems like Chickenfoot \citep{bolin2005} that
cannot be reproduced in Wildcard. We consider this an acceptable
tradeoff in exchange for achieving a gentler slope of customization, and
we show in Section~\ref{sec:examples} and Section~\ref{sec:evaluation}
that our model can still implement many useful customizations in
practice. Also, there is sometimes a way to reframe an imperative script
in terms of our direct manipulation model; for example, a script that
iterates through rows in a page adding some additional information could
be reproduced using a formula in Wildcard.

\hypertarget{third-party-semantic-wrappers}{%
\subsection{Third-party semantic
wrappers}\label{third-party-semantic-wrappers}}

@@

\emph{Third-party customization tools} enable customization without
using official extension APIs, enabling a broader range of
customizations on top of more applications. For example, web browser
extensions have demonstrated the utility of customizing websites through
manipulating the DOM, without needing explicit extension APIs to be
built in. However, third-party customization comes with a corresponding
cost: these tools can typically only operate at a low level of
abstraction, e.g.~manipulating user interface elements. This makes it
harder for end users to write scripts, and makes the resulting scripts
more brittle. (todo: support this claim more, provide examples)

\emph{Semantic customization tools} use explicit extension APIs provided
by the application developer. Examples of this include accessing a
backend web API, or writing a customization in Applescript that uses an
application-specific API. The main benefit is that this allows the
extension author to work with meaningful concepts in the application
domain---``create a new calendar event'' rather than ``click the button
that says new event.''---which makes customizations easier to build and
more robust. However, this style limits the range of extensions that can
be built, to only those which the official plugin API supports.

With Wildcard, we use a hybrid approach that takes the best of both
worlds. Programmers implement an API wrapper that is internally
implemented as a black-box customization, but externally provides a
semantic interface to the application. These wrappers are added to a
shared repository, available to all users of the system. When an end
user is using a site that already has an adapter, they get a semantic
customization experience that avoids low-level details.

One way to view this approach is as introducing a new abstraction
barrier into third-party extension. Typically, a third party
customization script combines two responsibilities: 1) mapping the
low-level details of a user interface to semantic constructs (e.g.,
using CSS selectors to find certain page elements), and 2) the actual
logic of the specific customization. Even though the mapping logic is
often more generic than the specific customization, the intertwining of
these two responsibilities in a single script makes it very difficult to
share the mapping logic across scripts.

With Wildcard we propose a decoupling of these two layers: a
community-maintained mapping layer, shared across many specific
customizations by individual users. This architecture has been
successfully used by projects like Gmail.js, an open source project that
creates a convenient API for browser extensions to interface with the
Gmail web email client.

@@

\begin{itemize}
\tightlist
\item
  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
\item
  browser dev tools
\item
  Self, Smalltalk
\item
  multiple representations
\item
  Semantic Web
\end{itemize}

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we show how the behavior of a software application can
  be extended and adapted by direct manipulation, using table-driven
  customization, a new paradigm that allows end users to customize
  applications without writing traditional code.

  Instead, users directly manipulate a tabular view of the structured
  data inside the application---rather than writing imperative scripts
  (as in most customization tools). This simple model also accommodates
  a spreadsheet formula language and custom data editing widgets, which
  provide sufficient expressivity to implement many useful
  customizations.

  We illustrate the approach with Wildcard, a browser extension that
  implements table-driven customization in the context of web
  applications. Through concrete examples, we show that our paradigm can
  be used to create useful customizations for real applications. We
  share reflections from experience using the Wildcard system, on both
  its strengths and limitations relative to other customization
  approaches. Finally, we explore how this paradigm might lead to new
  software architectures that encourage this form of customization.
\end{abstract}

@@

Most attempts at empowering end users to customize their software offer
a simplified version of programming. Some scripting languages
(AppleScript, Chickenfoot) have a friendly syntax that resembles natural
language. Visual programming tools (Mac Automator, Zapier) eliminate
text syntax entirely. Macro recorders (Applescript, Helena, WebVCR)
remove some of the initial programming burden by letting a user start
with concrete demonstrations.

Despite their many differences, these approaches all share something in
common: an imperative programming model, with statement sequencing,
mutable variables and loops. End users express their ideas in
scripts---sequences of commands---which, name aside, are not very
different from conventional code.

We have known for decades about an alternate approach: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the \emph{de facto} standard in GUIs today, but when it
comes to customizing software, it is rarely to be found. In this work,
we ask: what would it look like to build a software customization
interface that relies on direct manipulation? We take inspiration from
spreadsheets and visual database query interfaces
\citep{2020a, bakke2016}, which have successfully enabled end users to
run queries and computations through direct manipulation of data.

In this paper we present a technique called \emph{table-driven
customization}, which applies ideas from visual query interfaces in the
context of software customization. An application's UI is augmented with
a table view, where the user can see and manipulate the application's
internal data . Changes in the table view result in immediate
corresponding changes to the original user interface of the application,
enabling the user to customize an application with live feedback.

We have developed a browser extension called Wildcard which uses web
scraping techniques to implement table-driven customization for existing
Web applications. In Section~\ref{sec:examples}, we introduce the key
ideas of table-driven customization by presenting several examples of
real customizations implemented in Wildcard.

In Section~\ref{sec:architecture}, we explain the architecture of
table-driven customization. We focus on the \emph{table adapter}
abstraction, which allows many different types of underlying data to be
bidirectionally mapped to a table. We describe several types of table
adapters we've built in Wildcard, and also describe future adapters that
are supported by the general paradigm.

We have used Wildcard to build real customizations for 11 different
websites. In Section~\ref{sec:evaluation}, we present reflections from
this process, outlining the kinds of customizations we were able to
build, limitations we encountered, and reflections on the ease of
integrating scraping logic with real websites.

@@

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: End users should be able
  to customize an application by directly examining and modifying its
  data, rather than by writing imperative scripts.
\item
  \emph{Wrapping applications for customization}: Typically, tools that
  don't rely on official extension APIs resort to offering low-level
  APIs for customization. Instead, we propose a community-maintained
  library of semantic wrappers around existing applications, enabling
  end users to work with domain objects rather than low-level
  representations.
\end{itemize}

@@

First, the three different data sources are each bidirectionally mapped
to a table interface by a \textbf{table adapter}. The table adapter
defines how to map a particular type of data to a table, and what
effects edits should have on the original data source. In some cases,
the mapping logic is straightforward: the local storage adapter stores a
table of data, so the mapping to the table abstraction is trivial. In
other cases, the mapping is more involved: the DOM scraping adapter
implements web scraping logic to produce a table of data from the web
page, and turns edits to the table into DOM manipulations like
reordering rows of data on the page.

The three separate tables are then combined into a single table for the
end user to view and edit. The \textbf{query engine} is responsible for
creating this combined view, and routing the user's edits back to the
individual table adapters. In this example, the query engine has joined
the three tables together by a shared ID column, and sorted the result
by the name column.

@@

A key idea in table-driven customization is that a wide variety of data
sources can be mapped to the generic abstraction of an editable table.
In a relational database, the table matches the underlying storage
format, but in table-driven customization, the table is merely an
\emph{interface layer}. The data shown in the table is a projection of
some underlying state, and edits to the table can have complex effects
on the underlying state.

Externally, a table adapter must satisfy an abstract interface. The
first two parts of the interface resemble the interface exposed by a
table in a relational database:

\textbf{Returning a table}: A table adapter exposes a table of data: an
ordered list of records. Each record carries a unique identifier and
associates named attributes with values. Tables have a typed schema, so
the same attributes are shared across all records. A table adapter can
update the contents of a table at any time, in response to changes in
the underlying state (e.g., a DOM scraping adapter can update the table
when the page body changes). When data changes, the query view is
reactively updated in response.

\textbf{Making edits}: The query engine can issue a request to a table
adapter to make an edit to a record. The meaning of making an edit can
vary depending on the adapter: in the local storage adapter, an
annotation may be persisted into local storage; in the DOM scraping
adapter, an edit may represent filling in a form field. An adapter can
also mark values as read-only if it wouldn't be meaningful to edit them;
for example, the DOM scraping adapter typically marks page content as
read-only, except for editable form fields.

@@

Here we describe in more detail the table adapters that we have
implemented in Wildcard to power the customizations shown in
Section~\ref{sec:examples}.

\textbf{DOM scraping adapters} are the essential component that enables
Wildcard to interface with an existing website UI. In addition to the
standard web scraping problem of extracting a table of data from the
DOM, a scraping adapter must also manipulate the DOM to reorder rows,
edit form entries, and inject annotations as the table is edited.

Because each website has unique content, we rely on programmers to
create a DOM scraping adapter for each individual website to make it
available for customization in Wildcard. To make this approach viable,
we have built a library of functions that make it easy to implement a
scraping adapter, requiring the programmer only to implement the minimal
site-specific parts. The programmer writes a single Javascript function
which can use standard scraping techniques like CSS selectors and DOM
APIs to extract relevant elements from the page. The library functions
then wrap this function to implement the rest of the needed
functionality; for example, when the table is sorted, we find the DOM
elements corresponding to the rows in the table, remove them from the
page, and then reinsert them in the new order.

An \textbf{AJAX scraping adapter} intercepts AJAX requests made by a web
page, and extracts information from those requests. When available, this
tends to be a helpful technique because the data is already in a
structured form, and often includes information not shown in the UI. As
with DOM scraping adapters, we have made it easy for programmers to
create site-specific AJAX scraping adapters. A programmer writes a
function that specifies how to extract data from an AJAX request, and
the framework handles the details of intercepting requests and calling
the programmer-defined function.\footnote{So far we have only
  implemented AJAX scraping in the Firefox version of Wildcard, since
  Firefox has convenient APIs for intercepting requests. It appears
  possible to implement in Chrome as well, but we have not finished our
  implementation.}

The \textbf{local storage adapter} simply stores a table of data in the
browser. As shown in the example use cases, this can be useful for
persisting annotations in a private way, without uploading them to a web
service.

@@

We have designed the table adapter API to be general enough to support
other types of useful adapters in the future. Here are two such
possibilities:

\textbf{Integrated website adapters}: A key benefit of the table adapter
abstraction is that Wildcard is not coupled to web scraping as the only
means for integrating with existing sites, but can also accommodate
first party developers adding support directly into their own websites.
An ``integrated website adapter'' installed by the developer could
directly accesses the internal state of the application, providing the
same functionality as a DOM scraping adapter but in a more robust way.

With the advent of rich frontend web frameworks, structured application
state is now often available in the web client. We suspect it is
possible to create plugins for frontend frameworks that expose this
state to Wildcard with only minimal effort from the application
developers. To test this hypothesis, we have created an early prototype
of a Wildcard adapter for the Redux state management library, and used
it with the TodoMVC todo list application. To install the adapter, the
programmer only needs to make a few small changes to their app:

\emph{Exposing a table}: In Redux, app developers already represent the
state of a user interface as a single object. To expose a table of data,
the developer simply writes a function that maps the state object to a
table. \emph{Editing data values}: The developer defines how edits to
records in the table should map to edits in the state object.
\emph{Sorting/filtering the UI}: The developer inserts an additional
data processing step before the state object is rendered in the UI, so
that Wildcard can apply sorting and filtering. \emph{Annotations}: The
developer adds additional React components managed by Wildcard into
their UI tree, which render additional data as annotations.

In certain types of software, customizability is often touted as a key
selling point. An integrated website adapter would provide a way for
developers to integrate with an existing ecosystem of formulas and
customization tools, without needing to build all that functionality
from scratch.

\textbf{Shared storage adapter}: It would be useful to share user
annotations among users and across devices---for example,
collaboratively taking notes with friends on options for places to stay
on Airbnb. The existing Local Storage Adapter could be extended to share
live synchronized data with other users. This could be achieved through
a centralized web server, or through P2P connections that might provide
stronger privacy guarantees.

@@

Queries are processed in three steps. First, the query invokes a primary
DOM scraping table adapter that associates records in the result with
elements in the application's user interface. At minimum, the primary
table adapter needs to return record IDs and have the ability to
manipulate the application's UI. It can also optionally return data
about each record. Next, additional tables (AJAX data, local storage
data) are left-joined by ID. Finally, the result table can be sorted and
filtered by any column.

This query model can be viewed as a tiny (and rather constrained) subset
of the SQL query model. Despite its simplicity, this simple model has
proven sufficient for meeting the needs of customization in practice,
and minimizes the complexity of supporting more general and arbitrary
queries. But because it fits into the general paradigm of relational
queries, it could theoretically be extended to support a wider range of
queries.

The query engine is also responsible for executing formulas. We have
built a small formula language resembling the ones used in visual query
tools like SIEUFERD. As in those tools, and unlike in spreadsheets,
formulas automatically apply across an entire column of data, and
reference other column names instead of values on specific rows. For
example, the user would write \texttt{=price\ *\ 2} to create a column
with a value derived from a price attribute. This is more convenient
than needing to copy-paste a formula across an entire column as in
spreadsheets, and none of our customization use cases have uncovered a
need for writing different formulas on different rows of the table.

@@

Ainsworth et al provide a helpful taxonomy of the value of multiple
representations \citep{ainsworth1999}. In their terms, Wildcard plays a
\emph{complementary role} by supporting a different set of tasks from
the original application, while displaying \emph{shared information}. We
also suspect that Wildcard helps \emph{constructs deeper understanding}
by \emph{subtraction}. By stripping away details and only showing the
essential data in an interface, Wildcard helps users think of new ways
of using that data, outside the specific constraints of the original
application.

It is important to note that there are many specific customizations that
can be achieved in systems like Chickenfoot \citep{bolin2005} that
cannot be reproduced in Wildcard. We consider this an acceptable
tradeoff in exchange for achieving a gentler slope of customization, and
we show in Section~\ref{sec:examples} and Section~\ref{sec:evaluation}
that our model can still implement many useful customizations in
practice. Also, there is sometimes a way to reframe an imperative script
in terms of our direct manipulation model; for example, a script that
iterates through rows in a page adding some additional information could
be reproduced using a formula in Wildcard. (\emph{todo: mention the fact
that the data abstraction step limits possibilities})

\hypertarget{wrapping-applications-for-customization}{%
\subsection{Wrapping applications for
customization}\label{wrapping-applications-for-customization}}

@@

\emph{Ad hoc customization tools} enable customization without using
official extension APIs, enabling a broader range of customizations on
top of more applications. For example, web browser extensions have
demonstrated the utility of customizing websites through manipulating
the DOM, without needing explicit extension APIs to be built in.
However, ad hoc customization comes with a corresponding cost: these
tools can typically only operate at a low level of abstraction,
e.g.~manipulating user interface elements. This makes it harder for end
users to write scripts, and makes the resulting scripts more brittle.

\emph{Anticipated customization tools}, in contrast, use explicit
extension APIs provided by the application developer. Examples of this
include accessing a backend web API, or writing a customization in
Applescript that uses an application-specific API. The main benefit is
that this allows the extension author to work with meaningful concepts
in the application domain---``create a new calendar event'' rather than
``click the button that says new event.''---which makes customizations
easier to build and more robust. However, this style limits the range of
extensions that can be built to those that the official plugin API
supports.

With Wildcard, we use a hybrid approach that aims to provide the best of
both worlds. Programmers implement an API wrapper that is internally
implemented as an ad hoc customization, but externally provides a
high-level interface to the application, abstracting away the details of
the user interface. These wrappers are added to a shared repository,
available to all users of the system. When an end user is using a site
that already has an adapter, they get a semantic customization
experience that avoids low-level details.

One way to view this approach is as introducing a new abstraction
barrier into third-party extension. Typically, a third party
customization script combines two responsibilities: 1) mapping the
low-level details of a user interface to semantic constructs (e.g.,
using CSS selectors to find certain page elements), and 2) handling the
actual logic of the specific customization. Even though the mapping
logic is often more generic than the specific customization, the
intertwining of these two responsibilities in a single script makes it
very difficult to share the mapping logic across scripts.

With Wildcard we propose a decoupling of these two layers: a repository
of shared wrappers maintained by programmers, and a separate repository
of specific customizations built on top of these wrappers. This general
architecture has been successfully demonstrated by projects like
Gmail.js, an open source project that creates a convenient API for
browser extensions to interface with the Gmail web email client.

(\emph{todo: mention automated wrapper induction here or somewhere
else?})

@@

\begin{itemize}
\tightlist
\item
  web scraping / wrapper induction
\item
  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
\item
  browser dev tools
\item
  Self, Smalltalk
\item
  multiple representations
\item
  Semantic Web
\end{itemize}
